---
title: "Class 12: Transcriptomics and the analysis of RNA-Seq data"
author: Hannah Kim 
date: 05-12-23
format: pdf
---

# Import countData and colData

```{r}
counts <- read.csv("airway_scaledcounts.csv", row.names=1)
metadata <-  read.csv("airway_metadata.csv")
```

```{r}
head(counts)
head(metadata)

```

**Q1.** How many genes are in this dataset?

38694

**Q2.** How many 'control' cell lines do we have?

4

# Toy differential gene expression

```{r}
#Find the sample ids for control samples
control <- metadata[metadata[,"dex"]=="control",]
control.counts <- counts[ ,control$id]
control.mean <- rowSums( control.counts )/4 
head(control.mean)
```

```{r}
#Use these ids to select the columns I want in another table
metadata[metadata[,"dex"] == "control",]

control$id
```

```{r}
control.counts <- counts[ ,control$id]
control.mean <- rowSums( control.counts )/4 
head(control.mean)
```

**Q3.** How would you make the above code in either approach more robust?

Create a function that takes the "treated" or "control" data to calculate the mean so that the code is more simplified and reusable.

**Q4.** Follow the same procedure for the `treated` samples (i.e. calculate the mean per gene across drug treated samples and assign to a labeled vector called `treated.mean`)

```{r}
treated <- metadata[metadata[,"dex"]=="treated",]
treated.counts <- counts[ ,treated$id]
treated.mean <- rowSums( treated.counts )/4 
head(treated.mean)

```

**Q5 (a).** Create a scatter plot showing the mean of the treated samples against the mean of the control samples. Your plot should look something like the following.

```{r}
meancounts <- data.frame(control.mean, treated.mean)
colSums(meancounts)

# Plot both means by plotting 
plot(meancounts[,1],meancounts[,2], xlab="Control", ylab="Treated")

```

To calculate the log2 of the fold change between treated and control...

```{r}
meancounts$log2fc <- log2(meancounts$treated.mean/ meancounts$control.mean)
```

**Q7.** What is the purpose of the `arr.ind` argument in the **which()** function call above? Why would we then take the first column of the output and need to call the **unique()** function?

```{r}
zero.vals<- which(meancounts[, 1:2]==0, arr.ind =TRUE)
to.rm<- unique(zero.vals[,1])
mycounts<- meancounts[-to.rm]

```

The arr.ind is important because it returns the indexes of the values that are TRUE. We take the first column from this and return the non repeating or unique indexes.

**Q8.** Using the `up.ind` vector above can you determine how many up regulated genes we have at the greater than 2 fc level?

250

**Q9.** Using the `down.ind` vector above can you determine how many down regulated genes we have at the greater than 2 fc level?

367

**Q10.** Do you trust these results? Why or why not?

There is lots of documentation for existing functions such as the DESeq2 class that we can use to run statistical analysis/functions rather than doing it by hand and getting human errors.

# DESeq2 Analysis

First step, loading the library

```{r}
library('DESeq2')
```

```{r}
# Love, M.I., Huber, W., Anders, S. Moderated estimation of fold change and dispersion for RNA-seq data with DESeq2 Genome Biology 15(12):550 (2014)
citation('DESeq2')
```

Let's generate the specific object that DESeq2 needs:

```{r}
#Id needs to be Col Names. We use DEX to define the subgroups
dds <- DESeqDataSetFromMatrix(countData = counts, 
                              colData = metadata, 
                              design= ~dex)
dds

```

```{r}
dds <- DESeq(dds)
```

```{r}
#Show us the results of the statistical analysis 
res <- results(dds)
res
```

```{r}
summary(res)
```

```{r}
#Change the alpha from default of 0.1 to 0.05
res05 <- results(dds, alpha=0.05)
summary(res05)
```
